### 2017-04-10 Manuscripts/cog-psych/cogpsych.tex
added stats to jk13 analysis, adjusted binning

Because each experimenter-defined category in the \cite{jern2013probabilistic} experiment possessed a distinct hue, it is sensible to propose that participants would make their generated category distinct by using hues that differ from those present in the experimenter-defined category: If creative generation is influenced by category contrast, the hues of generated categories should be systematically different from those of the experimenter-defined categories. Problematically, however, stimulus hue was measured on an circular (HSV) scale with limits $[0,1]$: both 0 and 1 corresponded to a `red' hue, and all other possible hues were represented between these values. By consequence, proximity on the hue dimension does not directly correspond to perceptual similarity, and we cannot simply compare the hue values to determine the extent of the differences.

To overcome this issue, we categorized the hues of each participant's generated categories into one of six uniformly-spaced color groups: \{\textit{red}: 0 \& 1, \textit{yellow}: 0.167, \textit{green}: 0.333, \textit{teal}: 0.5, \textit{blue}: 0.667, \textit{pink}: 0.833\}. Each generated crystal was placed into the color group closest to its hue value. After each example's hue was categorized, we computed the amount of overlap between the hues of the six hues generated by each participant and the two hues present in the experimenter-defined categories: If contrast influences the hues of the generated categories, we should observe minimal overlap between the generated and experimenter-defined categories.

\nbcnote{Joe should probably do a pass through that language. Maybe add a cite for proximity-similarity distinction?}

\begin{figure}
    \begin{center}
        \inputpgf{figs/}{jk13-huecontrast.pgf}
        \caption{Analysis of data from \cite{jern2013probabilistic}, Experiment 3. Plotted is the number of generated items that share a color group with one of the experimenter-defined classes. The ``Expected'' data follows a Binomial distribution with $p=2/6$, given there were two experimenter-defined classes, and 6 color groups.}
        \label{fig:jk13-huecontrast}
    \end{center}
\end{figure}

These data, shown in Figure \ref{fig:jk13-huecontrast}, reveal a clear pattern: The majority of participants generated categories possessing entirely distinct hues; with 0/6 exemplars sharing a hue with the experimenter-defined categories. These results can be compared to the predictions of a Binomial model, which proposes that participants generate hues at random. According to the model, if hue selection is not systematic, the probability that any given example will lie in the same color group as an experimenter-defined category is $p = 2/6$, as there were two experimenter-defined categories and six possible color groups. Thus, the number of shared hues is Binomially distributed with $p=2/6$ and $n=6$ (as participants generated six examples in each condition). Comparison of the behavioral results to the model predictions reveals that participants were far more likely to select distinct hues: participants tended to generate items that were perceptually distinct from the categories they had learned.

These results are notable in that they provide some early, concrete evidence that creative generation is influenced by category contrast. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our results suggest that generated categories tend to be distinct from \textit{and} distributionally similar to what is already known. However, while these results indicate that contrast exerts \textit{some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.

@@

To begin our investigation, we sought to extend the early evidence from the \cite{jern2013probabilistic} data, under a variety of learning conditions and using more standard stimulus materials. We developed an artificial, two dimensional domain of squares, varying in color and size (see Figure \ref{fig:e1-conditions}, panel A). These features have been used in numerous classification learning studies \cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}. Unlike those used in the \cite{jern2013probabilistic} experiments, proximity along these dimensions aligns more directly with perceptual similarity, allowing us to more directly evaluate the role of category contrast in creative generation. In order to extend the evidence provided by the \cite{jern2013probabilistic} data, we began by testing effects of category contrast under a variety of learning conditions. This was achieved by training participants on categories possessing qualitatively different distributional structures, as shown in Figure \ref{fig:e1-conditions}.

@@

As in Experiment 1, we observed broad differences in the generation approach taken by different participants.  To characterize the nature of these differences, Figure \ref{fig:e2-samples} depicts sample categories generated by participants. The data from each condition in Figure \ref{fig:e2-samples} is organized into four columns based on commonly observed patterns of generation: a `Cluster' type of tightly-clustered examples,   `Row' and `Column' types of exemplars widely distributed along the one axis but narrowly along the other, and a `Corners' type, wherein participants placed exemplars in disparate corners of the space. As before, in this section we focus on what can be concluded from analyzing the data in aggregate, but in later sections we will focus more specifically on explaining the individual differences.

@@

\nbcnote{ Weak spot: the obvious next analysis here would involve the same box-and-whisker plot as in Experiment 1. We ran this analysis but the conditions did not differ in Y-range due to the data being weirdly distributed. But it feels odd to ignore this point, so I've included the box-and-whisker plot, as well as the graph-of-many-lines figure, on a separate page at the end.}

@@

\begin{figure}[p]
    \begin{center}
    \inputpgf{figs/}{range-diff-gradients.pgf}
    \caption{Behavioral and simulated relative range gradients. Each panel shows, for each stimulus, the dimensional orientation of the categories it was generated into: vertically aligned `columns' versus horizontally aligned `rows'.}
    \label{fig:range-diff-gradients}
    \end{center}
\end{figure}

>>>>

Because each experimenter-defined category in the \cite{jern2013probabilistic} experiment possessed a distinct hue, it is sensible to propose that participants would make their generated category distinct by using hues that differ from those present in the experimenter-defined category: If creative generation is influenced by category contrast, the hues of generated categories should be systematically different from those of the experimenter-defined categories. Problematically, however, stimulus hue was measured on an circular (HSV) scale with limits $[0,1]$: both 0 and 1 corresponded to a `red' hue, and all other possible hues were represented between these values. By consequence, proximity on the hue dimension does not directly correspond to perceptual similarity, and we cannot simply compare the hue values to determine the extent of the differences.
\nbcnote{Maybe add a cite for proximity-similarity distinction?}

To overcome this issue, we categorized all possible hues into one of eight uniformly-spaced color groups: \{\textit{Red}: $0-0.063$, $0.938-1$, \textit{Yellow}: $0.063-0.188$, \textit{Yellow-Green}: $0.188-0.313$, \textit{Green-Teal}: $0.313-0.438$, \textit{Teal}: $0.438-0.563$, \textit{Teal-Blue}: $0.563-0.688$, \textit{Purple}: $0.688-0.813$, \textit{Pink}: $0.813-0.938$\}. In the \cite{jern2013probabilistic} experiment, the hue of each experimenter-defined category was selected from one of six possible values, each of which falls into one of the color groups above. By categorizing the participant-generated crystals likewise, we can obtain a broad measure of category contrast by determining the proportion of participant-generated crystals that fall into the same groups as the experimenter-defined categories: If contrast influences the hues of the generated categories, we should observe minimal overlap between in the color groupings.

\begin{figure}
    \begin{center}
        \inputpgf{figs/}{jk13-huecontrast.pgf}
        \caption{Analysis of data from \cite{jern2013probabilistic}, Experiment 3. Plotted is the number of generated items that share a color group with one of the experimenter-defined classes. The ``Expected'' data follows a Binomial distribution with $p=2/8$, given there were two experimenter-defined classes, and eight color groups.}
        \label{fig:jk13-huecontrast}
    \end{center}
\end{figure}

These data, shown in Figure \ref{fig:jk13-huecontrast}, reveal a clear pattern: The majority of participants in each condition ($n = 22$) generated categories possessing entirely distinct hues; with 0/6 exemplars sharing a hue with the experimenter-defined categories. These results can be compared to the predictions of a Binomial model, which proposes that participants generate hues at random. That is, if hue selection is not systematic, the probability that any given example will lie in the same color group as an experimenter-defined category is given by a Binomial distribution with $p = 2/8$, as there were two experimenter-defined categories and eight possible color groups. Chi-square goodness-of-fit tests reveal that the observed distribution in each condition is highly inconsistent with the random-hue model (all $\tilde{\chi}^2>200$, $p<0.001$): Participants tended to generate items that were perceptually distinct from the categories they had learned, and were less likely to generate hues possessed by members of the experimenter-defined categories.

These results are notable in that they provide some early, concrete evidence that creative generation is influenced by category contrast. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our results suggest that generated categories tend to be distinct from \textit{and} distributionally similar to what is already known. However, it is worth noting that our analysis is still limited: the color groups defined above are somewhat arbitrary, and it is not clear that our color grouping is consistent with psychological color boundaries. While we did obtain similar results using a variety of alternative groupings, the hue dimension used in the \cite{jern2013probabilistic} study does not lend itself straightforwardly to the the computation of similarities, and thus we cannot be certain of whether our coding accurately approximates the psychological space of the stimuli. By consequence, although these results likely indicate that contrast exerts \textit{some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.

\nbcnote{could use a cite on why hue dimension proximity does not match up with similarity. maybe also a note about how we can't be sure what hues were actually perceived due to norming, etc.}

@@

To begin our investigation, we sought to extend the early evidence obtained from our analysis of the \cite{jern2013probabilistic} data, under a variety of learning conditions and using more standard stimulus materials. We developed an artificial, two dimensional domain of squares, varying in color and size (see Figure \ref{fig:e1-conditions}, panel A). These features have been used in numerous classification learning studies \cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}. Unlike those used in the \cite{jern2013probabilistic} experiments, proximity along these dimensions aligns more directly with perceptual similarity, allowing us to more directly evaluate the role of category contrast in creative generation. In order to extend the evidence provided by the \cite{jern2013probabilistic} data, we began by testing effects of category contrast under a variety of learning conditions. This was achieved by training participants on categories possessing qualitatively different distributional structures, as shown in Figure \ref{fig:e1-conditions}.

@@

As in Experiment 1, we observed broad differences in the generation approach taken by different participants.  To characterize the nature of these differences, Figure \ref{fig:e2-samples} depicts sample categories generated by participants. The data from each condition is organized into four columns based on commonly observed patterns of generation: a `Cluster' type of tightly-clustered examples,   `Row' and `Column' types of exemplars widely distributed along the one axis but narrowly along the other, and a `Corners' type, wherein participants placed exemplars in disparate corners of the space. As before, in this section we focus on what can be concluded from analyzing the data in aggregate, but in later sections we will focus more specifically on explaining the individual differences.

@@

\nbcnote{ Weak spot: the obvious next analysis here would involve the same box-and-whisker plot as in Experiment 1. We ran this analysis but the conditions did not differ in Y-range due to the data being weirdly distributed. I've included the box-and-whisker plot, as well as the graph-of-many-lines figure, on a separate page at the end.}

@@

\begin{figure}[p]
    \begin{center}
    \inputpgf{figs/}{range-diff-gradients.pgf}
    \caption{Behavioral and simulated relative range gradients. Each panel shows, for each stimulus, the dimensional orientation of the categories it was generated into: vertically aligned `columns' versus horizontally aligned `rows'.}
    \nbcnote{Need to run this on lando with many more samples}
    \label{fig:range-diff-gradients}
    \end{center}
\end{figure}
